Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-06-22 optim.tex
squintability subsection
added the content of the squintability subsection, with equations

\affiliation[1]{organization={University of Texas at Austin, Department
of Statistics and Data Sciences},city={Austin},country={United
States},countrysep={,},postcode={78751},postcodesep={}}
\affiliation[2]{organization={Monash University, Department of
Econometrics and Business
Statistics},city={Melbourne},country={Australia},countrysep={,},postcode={3800},postcodesep={}}
\affiliation[3]{organization={BNU-HKBU United International
College, Department of Mathematical
Sciences},city={Zhuhai},country={China},countrysep={,},postcode={519087},postcodesep={}}

@@

\begin{keyword}
    projection pursuit \sep Jellyfish Search Optimiser (JSO) \sep 
    Optimisation
\end{keyword}
\end{frontmatter}

@@

\section{Introduction {[}Nicolas and
Jessica{]}}\label{introduction-nicolas-and-jessica}

Projection Pursuit (PP) is a high-dimensional data visualisation
approach that involves computing informative linear projections of data
by optimising a variety of objective functions, namely \emph{PP index}
(\citet{hall1989polynomial}, \citet{cook1993projection},
\citet{lee2010projection}, \citet{Loperfido2018},
\citet{Loperfido2020}). Given high-dimensional data
\(X \in \mathbf{R}^{n }\times p\) and the index function \(f(\cdot)\),
PP finds the orthonormal projection matrix
\(A \in \mathbf{R}^{p \times d}\) by solving the following optimisation
problem:

\[
\underset{A}{\max } \quad f(XA) \quad \text{subject to} \quad A'A = I
\]

@@

The primary goal of this study is to investigate the performance of JSO
in the context of the Projection pursuit. A series of simulation
experiments using well-recognised data sets and PP indexes are conducted
to yield insights regarding the behaviour of JSO under different
settings and the sensitivity of hyper-parameter in the optimisation.
Second, to observe the performance of JSO with different types of PP
indexes, a collection of metrics is introduced to capture specific
properties of the index including squintability and smoothness as
introduced in \citet{laa_using_2020}. To the best of the authors'
knowledge, this is the first attempt to quantitatively measure
squintability and smoothness. Finally, the relationship between the JSO
performance, hyper-parameters tuning and various properties of PP
indexes is analysed to provide helpful guidance for practitioners that
are using the guided tour for high dimensional visualisation.

@@

An initial investigation of PP indexes, and the potential for
scagnostics is described in \citet{laa_using_2020}. To be useful here an
optimiser needs to be able to handle functions which are not very
smooth. In addition, because data structures might be relatively fine,
the optimiser needs to be able to find maxima that occur with a small
squint angle, that can only be seen from very close by. One last aspect
that is useful is for an optimiser to return local maxima in addition to
global because data can contain many different and interesting features.

\section{The jellyfish optimiser and properties of PP indexes {[}Nicolas
and Jessica{]}}\label{sec-theory}

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, toptitle=1mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, left=2mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, arc=.35mm, opacityback=0, breakable, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, bottomtitle=1mm, titlerule=0mm, toprule=.15mm, leftrule=.75mm, colback=white]

@@

In this study, we propose two metrics to capture the notion of
squintability.

@@

First, parametric model.

{[}Nicolas's pdf{]}

Second, we consider the product of the largest absolute magnitude of
rate of change of \(f\) and the corresponding projection angle as a
second measure of squintability. Since \(f\) is decreasing, the rate of
change of \(f\) is negative and thus \(|\underset{x}{\min} f('x)|\)
gives the absolute magnitude of the most negative rate of change.

{[}Nicolas's pdf{]}

@@

\section{Visualisation of jellyfish optimiser}\label{sec-vis}

@@

\subsection{Performance of JSO in pipe-finding
problems}\label{sec-app-1}

@@

\subsection{Factors affecting JSO success rate: optimisation properties
and jellyfish hyper-parameters}\label{sec-app-2}

@@

\begin{tabular}{|>{}lll>{}r|rrr>{}r|rrrr>{}r|}
\toprule
  & shape & index & d & variance & range & smooth & nugget & $\theta_1$ & $\theta_2$ & $\theta_3$ & $\theta_4$ & squint\\
\midrule
1 & pipe & holes & 6 & 0.002 & 0.371 & 2.355 & 0.221 & 1.015 & 0.860 & 3.368 & 0.018 & 3.053\\
2 & pipe & holes & 8 & 0.000 & 0.178 & 2.189 & 0.820 & 1.014 & 0.869 & 3.264 & 0.029 & 2.960\\
3 & pipe & holes & 10 & 0.000 & 0.107 & 2.192 & 3.027 & 1.016 & 0.885 & 3.151 & 0.022 & 2.948\\
4 & pipe & holes & 12 & 0.000 & 0.148 & 2.295 & 1.581 & 1.011 & 0.878 & 3.345 & 0.004 & 3.117\\
5 & sine & MIC & 6 & 0.016 & 0.092 & 2.457 & 0.083 & 0.894 & 0.571 & 1.623 & -0.024 & 1.257\\
\addlinespace
6 & sine & MIC & 8 & 0.016 & 0.081 & 2.636 & 0.084 & 0.932 & 0.328 & 1.314 & -0.030 & 0.772\\
7 & sine & TIC & 6 & 0.124 & 0.110 & 2.444 & 0.086 & 0.951 & 0.536 & 1.719 & -0.027 & 1.322\\
8 & sine & TIC & 8 & 0.125 & 0.099 & 2.475 & 0.085 & 0.949 & 0.564 & 1.723 & -0.028 & 1.369\\
9 & sine & dcor2d & 6 & 0.034 & 0.167 & 2.663 & 0.114 & 0.954 & 1.039 & 2.742 & -0.019 & 2.947\\
10 & sine & loess2d & 6 & 0.079 & 0.336 & 1.988 & 0.309 & 1.016 & 1.039 & 2.648 & 0.080 & 2.756\\
\addlinespace
11 & sine & splines2d & 6 & 0.042 & 0.242 & 2.537 & 0.102 & 1.014 & 1.051 & 2.730 & -0.009 & 3.118\\
12 & sine & stringy & 6 & 0.000 & 0.007 & 1.536 & 38.166 & 1.045 & 0.011 & 254.734 & 0.053 & 2.957\\
\bottomrule
\end{tabular}
\endgroup{}

\end{table}%

@@

\begin{longtable}[t]{lrrrrrrl}

\caption{\label{tbl-mod-data}The first 7 rows of the datasets processed
for modelling.}

\tabularnewline

\toprule
Index & D & Smoothness & Squintability & Num. of jellyfish & Max. Num. of tries & Prop. of success & Time\\
\midrule
MIC & 6 & 2.457 & 1.257 & 20 & 50 & 0.12 & 2.479 secs\\
MIC & 6 & 2.457 & 1.257 & 20 & 100 & 0.24 & 8.950 secs\\
MIC & 6 & 2.457 & 1.257 & 50 & 50 & 0.52 & 5.651 secs\\
MIC & 6 & 2.457 & 1.257 & 50 & 100 & 0.64 & 13.223 secs\\
MIC & 6 & 2.457 & 1.257 & 100 & 50 & 0.76 & 19.453 secs\\
\addlinespace
MIC & 8 & 2.636 & 0.772 & 20 & 50 & 0.08 & 2.566 secs\\
MIC & 8 & 2.636 & 0.772 & 20 & 100 & 0.08 & 4.960 secs\\
\bottomrule

\end{longtable}

\endgroup{}

@@

\begin{longtable}[t]{|>{}lrrr>{}r|}

\caption{\label{tbl-mod-output}Model estimates of proportion of
jellyfish success on optimisation properties and jellyfish
hyper-parameters from the generalised linear model with a binomial
family and a logit link function. The variable smoothness,
squintability, number of jellyfish and maximum number of tries are
positively associated with JSO success rate while data dimension and
being flagged as long runtime are negatively associated with the success
rate. The variable squintability and dimension are significant,
suggesting their importance relative to jellyfish hyper-parameters in
the optimisation success.}

\tabularnewline

\toprule
term & estimate & std.error & statistic & p.value\\
\midrule
Intercept & -4.523 & 5.332 & -0.848 & 0.396\\
Smoothness & 1.195 & 1.915 & 0.624 & 0.533\\
Squintability & 2.060 & 0.684 & 3.011 & 0.003\\
dimension (d) & -0.628 & 0.255 & -2.462 & 0.014\\
long time & -0.874 & 1.286 & -0.680 & 0.497\\
\addlinespace
number of jellyfish & 0.216 & 0.127 & 1.701 & 0.089\\
maximum number of tries & 0.113 & 0.150 & 0.752 & 0.452\\
\bottomrule

\end{longtable}

\endgroup{}

\section{Conclusion {[}Di and Sherry{]}}\label{sec-conclusion}

>>>>

\affiliation[1]{organization={University of Texas at Austin, Department
of Statistics and Data Sciences},city={Austin},country={United
States},countrysep={,},postcode={78751},postcodesep={}}
\affiliation[2]{organization={Monash University, Department of
Econometrics and Business
Statistics},city={Melbourne},country={Australia},countrysep={,},postcode={3800},postcodesep={}}
\affiliation[3]{organization={Guangdong Provincial Key Laboratory of
Interdisciplinary Research and Application for Data Science, BNU-HKBU
United International College, Department of Mathematical
Sciences},city={Zhuhai},country={China},countrysep={,},postcode={519087},postcodesep={}}

@@

\begin{keyword}
    projection pursuit \sep Jellyfish Search Optimiser (JSO) \sep 
    Optimisation
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden]}{\end{tcolorbox}}\fi

@@

\hypertarget{introduction-nicolas-and-jessica}{%
\section{Introduction {[}Nicolas and
Jessica{]}}\label{introduction-nicolas-and-jessica}}

Projection Pursuit (PP) is a high-dimensional data visualisation
approach that involves computing informative linear projections of data
by optimising a variety of objective functions, namely \emph{PP index}
(\citet{hall1989polynomial}, \citet{cook1993projection},
\citet{lee2010projection}, \citet{Loperfido2018},
\citet{Loperfido2020}). Given high-dimensional data
\(X \in \mathbf{R}^{n\times p}\) and the index function \(f(\cdot)\), PP
finds the orthonormal projection matrix
\(A \in \mathbf{R}^{p \times d}\) by solving the following optimisation
problem:

\begin{equation}\protect\hypertarget{eq-optimization}{}{
\underset{A}{\max } \quad f(XA) \quad \text{subject to} \quad A'A = I
}\label{eq-optimization}\end{equation}

@@

The primary goal of this study is to investigate the performance of JSO
in the context of the Projection pursuit. A series of simulation
experiments using well-recognised data sets and PP indexes are conducted
to yield insights regarding the behaviour of JSO under different
settings and the sensitivity of hyper-parameter in the optimisation.
Second, to observe the performance of JSO with different types of PP
indexes, a collection of metrics is introduced to capture specific
properties of the index including squintability and smoothness as
introduced in \citet{laa_using_2020}. To the best of the authors'
knowledge, this is the first attempt to quantitatively measure
squintability and smoothness in this context. Finally, the relationship
between the JSO performance, hyper-parameters tuning and various
properties of PP indexes is analysed to provide helpful guidance for
practitioners that are using the guided tour for high dimensional
visualisation.

@@

An initial investigation of PP indexes, and the potential for
scagnostics is described in \citet{laa_using_2020}. To be useful here an
optimiser needs to be able to handle index functions which are possibly
not very smooth. In addition, because data structures might be
relatively fine, the optimiser needs to be able to find maxima that
occur with a small squint angle, that can only be seen from very close
by. One last aspect that is useful is for an optimiser to return local
maxima in addition to global because data can contain many different and
interesting features.

\hypertarget{sec-theory}{%
\section{The jellyfish optimiser and properties of PP indexes {[}Nicolas
and Jessica{]}}\label{sec-theory}}

@@

\begin{tcolorbox}[enhanced jigsaw, opacityback=0, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, titlerule=0mm, bottomrule=.15mm, toprule=.15mm, leftrule=.75mm, bottomtitle=1mm, breakable, toptitle=1mm, arc=.35mm, colback=white, rightrule=.15mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, left=2mm, coltitle=black, colbacktitle=quarto-callout-note-color!10!white]

@@

In this subsection, a metric to capture the notion of squintability is
proposed, and two approaches to compute this metric numerically are
detailed.

\textcolor{orange}{Here we need to add a precise, mathematical
definition of the $x$ and $y$ variables ("angle distance" to optimal?
and PP index values) used in the figures of this Section.}

@@

Let \(f\) represent the average PP index value with respect to the
projection angle distance to optimal, over the course of the JSO. It is
expected that for a PP index with high squint angle, the optimization
(\ref{eq-optimization}) should make substantial progress early on.
Conversely, for a PP index with low squint angle, it might take a long
while for the optimization to make substantial progress, as the
candidate projections would need to be very close to the optimal one for
the structure of the index function to be visible enough to be amenable
to efficient optimization. This observation suggests that the extreme
values of \(f'\) (the ones for which \(f''=0\), assuming that \(f\) is
twice differentiable), and the angles for which these values are
attained, should play an important role in the mathematical definition
of squintability. Noticing that \(f\) is expected to be a decreasing
function, we propose the following squintability metric:

\begin{equation}\protect\hypertarget{eq-squintability}{}{
\varsigma:=-4\min_{x}f'(x)\times\underset{x}{\arg\min}f'(x)
}\label{eq-squintability}\end{equation}

On the one hand, \(-\min_{x}f'(x)\) represents the largest value of the
gradient of \(-f\). On the other hand, \(\underset{x}{\arg\min}f'(x)\)
represents the angle value at which this largest gradient is attained.
Based on the above discussion, it is expected that these two values
should be both high in the case of high squintability (fast increase in
\(f\) early on), and both low in the case of low squintability (any
substantial increase in \(f\) happens very late, close to the optimal
angle). This suggest that their product (\ref{eq-squintability}) should
provide a sensible measure of squintability. The multiplicative constant
\(4\), which can be deemed arbitrary, does not change the interpretation
of the squintability metric \(\varsigma\); it is here to adjust the
range of values of \(\varsigma\) and simplify the explicit formula for
\(\varsigma\) obtained later on.

To compute the squintability metric (\ref{eq-squintability}) in
practice, several approaches are possible. The first one is to propose a
parametric model for \(f\), and use it to obtain an explicit formula for
\(\varsigma\). Numerical experiments suggest a scaled sigmoid shape as
described below. Define

\begin{equation}\protect\hypertarget{eq-logistic}{}{
\ell(x):=\frac{1}{1+\exp(\theta_{3}(x-\theta_{2}))}\ ,
}\label{eq-logistic}\end{equation}

which is a decreasing logistic function depending on two parameters
\(\theta_2\) and \(\theta_3\) , such that
\(\ell(\theta_{2})=\frac{1}{2}\). Then, define

\begin{equation}\protect\hypertarget{eq-parametric}{}{
f(x)=(\theta_{1}-\theta_{4})\frac{\ell(x)-\ell(x_{\max})}{\ell(0)-\ell(x_{\max})}+\theta_{4}\ ,
}\label{eq-parametric}\end{equation}

which depends on three additional parameters, \(\theta_1\), \(\theta_2\)
, and \(x_{\max}\), such that \(f(0)=\theta_1\) and
\(f(x_{\max})=\theta_4\). Under the parametric model
(\ref{eq-parametric}), the squintability metric (\ref{eq-squintability})
can be shown to be equal to

\begin{equation}\protect\hypertarget{eq-squintability-parametric}{}{
\varsigma=\frac{(\theta_{1}-\theta_{4})\theta_{2}\theta_{3}}{\ell(0)-\ell(x_{\max})}
}\label{eq-squintability-parametric}\end{equation}

In practice, the parameters of this model (\ref{eq-parametric}) can be
estimated numerically, for example by non-linear least squares, and then
used to evaluate \(\varsigma\) as in equation
(\ref{eq-squintability-parametric}).

Alternatively, one can estimate (\ref{eq-squintability}) in a
nonparametric way, for example by fitting \(f\) using kernel regression,
then numerically estimate the angle at which \(-f'\) attains its highest
value.

\textcolor{orange}{See whether further details are needed for the nonparametric approach.}

@@

\hypertarget{sec-vis}{%
\section{Visualisation of jellyfish optimiser}\label{sec-vis}}

@@

\hypertarget{sec-app-1}{%
\subsection{Performance of JSO in pipe-finding
problems}\label{sec-app-1}}

@@

\hypertarget{sec-app-2}{%
\subsection{Factors affecting JSO success rate: optimisation properties
and jellyfish hyper-parameters}\label{sec-app-2}}

@@

\begin{tabular}{|>{}lll>{}r|rrr>{}r|rrrr>{}r|}
\toprule
  & shape & index & d & variance & range & smooth & nugget & $\theta_1$ & $\theta_2$ & $\theta_3$ & $\theta_4$ & squint\\
\midrule
1 & pipe & holes & 6 & 0.002 & 0.371 & 2.355 & 0.221 & 1.015 & 0.860 & 3.368 & 0.018 & 3.053\\
2 & pipe & holes & 8 & 0.000 & 0.178 & 2.189 & 0.820 & 1.014 & 0.869 & 3.264 & 0.029 & 2.960\\
3 & pipe & holes & 10 & 0.000 & 0.107 & 2.192 & 3.027 & 1.016 & 0.885 & 3.151 & 0.022 & 2.948\\
4 & pipe & holes & 12 & 0.000 & 0.148 & 2.295 & 1.581 & 1.011 & 0.878 & 3.345 & 0.004 & 3.117\\
5 & sine & MIC & 6 & 0.016 & 0.092 & 2.457 & 0.083 & 0.894 & 0.571 & 1.623 & -0.024 & 1.257\\
\addlinespace
6 & sine & MIC & 8 & 0.016 & 0.081 & 2.636 & 0.084 & 0.932 & 0.328 & 1.314 & -0.030 & 0.772\\
7 & sine & TIC & 6 & 0.124 & 0.110 & 2.444 & 0.086 & 0.951 & 0.536 & 1.719 & -0.027 & 1.322\\
8 & sine & TIC & 8 & 0.125 & 0.099 & 2.475 & 0.085 & 0.949 & 0.564 & 1.723 & -0.028 & 1.369\\
9 & sine & dcor2d & 6 & 0.034 & 0.167 & 2.663 & 0.114 & 0.954 & 1.039 & 2.742 & -0.019 & 2.947\\
10 & sine & loess2d & 6 & 0.079 & 0.336 & 1.988 & 0.309 & 1.016 & 1.039 & 2.648 & 0.080 & 2.756\\
\addlinespace
11 & sine & splines2d & 6 & 0.042 & 0.242 & 2.537 & 0.102 & 1.014 & 1.051 & 2.730 & -0.009 & 3.118\\
12 & sine & stringy & 6 & 0.000 & 0.007 & 1.536 & 38.166 & 1.045 & 0.011 & 254.734 & 0.053 & 2.957\\
\bottomrule
\end{tabular}
\endgroup{}
\end{table}

@@

\hypertarget{tbl-mod-data}{}
\begin{longtable}[t]{lrrrrrrl}
\caption{\label{tbl-mod-data}The first 7 rows of the datasets processed for modelling. }\tabularnewline

\toprule
Index & D & Smoothness & Squintability & Num. of jellyfish & Max. Num. of tries & Prop. of success & Time\\
\midrule
MIC & 6 & 2.457 & 1.257 & 20 & 50 & 0.12 & 2.479 secs\\
MIC & 6 & 2.457 & 1.257 & 20 & 100 & 0.24 & 8.950 secs\\
MIC & 6 & 2.457 & 1.257 & 50 & 50 & 0.52 & 5.651 secs\\
MIC & 6 & 2.457 & 1.257 & 50 & 100 & 0.64 & 13.223 secs\\
MIC & 6 & 2.457 & 1.257 & 100 & 50 & 0.76 & 19.453 secs\\
\addlinespace
MIC & 8 & 2.636 & 0.772 & 20 & 50 & 0.08 & 2.566 secs\\
MIC & 8 & 2.636 & 0.772 & 20 & 100 & 0.08 & 4.960 secs\\
\bottomrule
\end{longtable}
\endgroup{}

@@

\hypertarget{tbl-mod-output}{}
\begin{longtable}[t]{|>{}lrrr>{}r|}
\caption{\label{tbl-mod-output}Model estimates of proportion of jellyfish success on optimisation
properties and jellyfish hyper-parameters from the generalised linear
model with a binomial family and a logit link function. The variable
smoothness, squintability, number of jellyfish and maximum number of
tries are positively associated with JSO success rate while data
dimension and being flagged as long runtime are negatively associated
with the success rate. The variable squintability and dimension are
significant, suggesting their importance relative to jellyfish
hyper-parameters in the optimisation success. }\tabularnewline

\toprule
term & estimate & std.error & statistic & p.value\\
\midrule
Intercept & -4.523 & 5.332 & -0.848 & 0.396\\
Smoothness & 1.195 & 1.915 & 0.624 & 0.533\\
Squintability & 2.060 & 0.684 & 3.011 & 0.003\\
dimension (d) & -0.628 & 0.255 & -2.462 & 0.014\\
long time & -0.874 & 1.286 & -0.680 & 0.497\\
\addlinespace
number of jellyfish & 0.216 & 0.127 & 1.701 & 0.089\\
maximum number of tries & 0.113 & 0.150 & 0.752 & 0.452\\
\bottomrule
\end{longtable}
\endgroup{}

\hypertarget{sec-conclusion}{%
\section{Conclusion {[}Di and Sherry{]}}\label{sec-conclusion}}
